---
title: "BMS501 - Case mapping"
author: "Siobhon Egan"
date: "18 March 2026"
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 4 # Adjust this number to include all header levels you need
---

Spatial aspects of your data can provide a lot of insights into the situation of the outbreak, and to answer questions such as:

-   Where are the current disease hotspots?
-   How have the hotspots have changed over time?
-   How is the access to health facilities? Are any improvements needed?

We will explore basic spatial data visualization methods using `tmap` and `ggplot2` packages. We will also walk through some of the basic spatial data management and querying methods with the sf package.

## Install and load packages

You will need to install the package pacman the first time run you run this. You can do this by executing the chunk below. Using `p_load` command from the `pacman` package makes it easier to load/install libraries.

```{r, install}
#install.packages("pacman")
#install.packages("rmarkdown")
```

Then we use the `p_load()` command from this package - it is a handy way to load packages - because it first checks if you have installed the package and if not it attempts to install it automatically for you from CRAN and/or any other repositories supported in the package repository list.

```{r, pload}
pacman::p_load(
  rio, # to import data
  tidyverse, # to clean, handle, and plot the data (includes ggplot2 package)
  sf, # to manage spatial data using a Simple Feature format
  tmap, # to produce simple maps, works for both interactive and static maps
  janitor # examining and cleaning data
)
```

## Sample case data

The case data is from the example in the epirhandbook https://epirhandbook.com/en/new_pages/gis.html#sample-case-data.
You can download a `.csv` version of the data from [here](https://github.com/siobhon-egan/BMS501-case-mapping/blob/main/data/linelist_cleaned.csv) if you haven't already. Ensure it is in a directory called **`data`**

For demonstration purposes, we will work with a random sample of 1000 cases from the simulated Ebola epidemic `linelist` dataframe (computationally, working with fewer cases is easier to display in this handbook).

This code assumes you set up your folder structure following the instructions on the LMS workshop 4 set up page. All your data files should be in the **data/** directory. Otherwise you can access all the files [here](https://github.com/siobhon-egan/BMS501-case-mapping/blob/main/data/).

Since we are taking a random sample of the cases, your results may look slightly different from what is demonstrated here when you run the codes on your own.

```{r, import-data}
# import clean case linelist
linelist <- import("data/linelist_cleaned.csv")
```

Next we select a random sample of 1000 rows using `sample()` from base R.

```{r, subsample}
# generate 1000 random row numbers, from the number of rows in linelist
sample_rows <- sample(nrow(linelist), 1000)

# subset linelist to keep only the sample rows, and all columns
linelist <- linelist[sample_rows,]
```

Now we want to convert this `linelist` which is class dataframe, to an object of class "sf" (spatial features). Given that the linelist has two columns "lon" and "lat" representing the longitude and latitude of each case's residence, this will be easy.

We use the package **sf** (spatial features) and its function `st_as_sf()` to create the new object we call `linelist_sf`. This new object looks essentially the same as the linelist, but the columns `lon` and `lat` have been designated as coordinate columns, and a coordinate reference system (CRS) has been assigned for when the points are displayed. 4326 identifies our coordinates as based on the [World Geodetic System 1984 (WGS84)](https://gisgeography.com/wgs84-world-geodetic-system/) - which is standard for GPS coordinates.

```{r, create-sf}
# Create sf object
linelist_sf <- linelist %>%
     st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

This is how the original `linelist` dataframe looks like. In this demonstration, we will only use the column `date_onset` and `geometry` (which was constructed from the longitude and latitude fields above and is the last column in the data frame).

To import a shapefile we use the `read_sf()` function from **sf**. If you set this up following the LMS instructions you should already have the file within the project directory **data/sle_adm3** directory with filename `sle_adm3.shp`. Otherwise you can load the zipped folder directly from [eppliedEpi repository here](https://github.com/appliedepi/appliedepidata/blob/main/inst/extdata/sle_adm3.zip)

```{r, read-shp, echo=F}
sle_adm3_raw <- read_sf("data/sle_adm3/sle_adm3.shp")
```

Next we use `clean_names()` from the **janitor** package to standardize the column names of the shapefile. We also use `filter()` to keep only the rows with admin2name of "Western Area Urban" or "Western Area Rural".

```{r, clean-shp}
# ADM3 level clean
sle_adm3 <- sle_adm3_raw %>%
  clean_names() %>% # standardize column names
  filter(admin2name %in% c("Western Area Urban", "Western Area Rural")) # filter to keep certain areas
```

Below you can see the how the shapefile looks after import and cleaning. *Scroll to the right* to see how there are columns with admin level 0 (country), admin level 1, admin level 2, and finally admin level 3. Each level has a character name and a unique identifier "pcode". The pcode expands with each increasing admin level e.g. SL (Sierra Leone) -\> SL04 (Western) -\> SL0410 (Western Area Rural) -\> SL040101 (Koya Rural).

### Population data 

**Sierra Leone: Population by ADM3**

Next we want to load in some population data - once again this is already available for you in the **data/** directory. Or the data can be downloaded from HDX [here](https://data.humdata.org/dataset/sierra-leone-population) . We use `import()` to load the .csv file. We also pass the imported file to `clean_names()` to standardize the column name syntax.

```{r}
# Population by ADM3
sle_adm3_pop <- import("data/sle_admpop_adm3_2020.csv") %>%
  clean_names()
```

Here is what the population file looks like. Scroll to the right to see how each jurisdiction has columns with `male` population, `female` populaton, `total` population, and the population break-down in columns by age group.

### Health Facilities 

**Sierra Leone: Health facility data from OpenStreetMap**

Next we load in another shape file related to location of health facilities - already downloaded in the **data/sle_adm3** directory with filename "sle_hf.shp".

We import the facility points shapefile with `read_sf()`, again clean the column names, and then filter to keep only the points tagged as either "hospital", "clinic", or "doctors".

```{r}
# OSM health facility shapefile
sle_hf <- read_sf("data/sle_adm3/sle_hf.shp") %>% 
  clean_names() %>%
  filter(amenity %in% c("hospital", "clinic", "doctors"))
```

Here is the resulting dataframe - *scroll right* to see the facility name and `geometry` coordinates.

<!-- ======================================================= -->

## Plotting coordinates

The easiest way to plot X-Y coordinates (longitude/latitude, points), in this case of cases, is to draw them as points directly from the `linelist_sf` object which we created in the preparation section.

The package **tmap** offers simple mapping capabilities for both static ("plot" mode) and interactive ("view" mode) with just a few lines of code. The **tmap** syntax is similar to that of **ggplot2**, such that commands are added to each other with `+`. Read more detail in this [vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html).

1)  Set the **tmap** mode. In this case we will use "plot" mode, which produces static outputs.

```{r, warning = F, message=F}
tmap_mode("plot") # choose either "view" or "plot"
```

Below, the points are plotted alone.`tm_shape()` is provided with the `linelist_sf` objects. We then add points via `tm_dots()`, specifying the size and color. Because `linelist_sf` is an sf object, we have already designated the two columns that contain the lat/long coordinates and the coordinate reference system (CRS):

```{r, warning = F, message=F}
# Just the cases (points)
tm_shape(linelist_sf) + tm_dots(size=0.08, col='blue')
```

Alone, the points do not tell us much. So we should also map the administrative boundaries:

Again we use `tm_shape()` (see [documentation](https://www.rdocumentation.org/packages/tmap/versions/3.3/topics/tm_shape)) but instead of providing the case points shapefile, we provide the administrative boundary shapefile (polygons).

With the `bbox =` argument (bbox stands for "bounding box") we can specify the coordinate boundaries. First we show the map display without `bbox`, and then with it.

```{r, out.width = c('50%', '50%'), fig.show='hold', warning=F, message=F}
# Just the administrative boundaries (polygons)
tm_shape(sle_adm3) +               # admin boundaries shapefile
  tm_polygons(col = "#F7F7F7")+    # show polygons in light grey
  tm_borders(col = "#000000",      # show borders with color and line weight
             lwd = 2) +
  tm_text("admin3name")            # column text to display for each polygon


# Same as above, but with zoom from bounding box
tm_shape(sle_adm3,
         bbox = c(-13.3, 8.43,    # corner
                  -13.2, 8.5)) +  # corner
  tm_polygons(col = "#F7F7F7") +
  tm_borders(col = "#000000", lwd = 2) +
  tm_text("admin3name")

```

And now both points and polygons together:

```{r, warning=F, message=FALSE}
# All together
tm_shape(sle_adm3, bbox = c(-13.3, 8.43, -13.2, 8.5)) +     #
  tm_polygons(col = "#F7F7F7") +
  tm_borders(col = "#000000", lwd = 2) +
  tm_text("admin3name")+
tm_shape(linelist_sf) +
  tm_dots(size=0.08, col='blue', alpha = 0.5) +
  tm_layout(title = "Distribution of Ebola cases")   # give title to map

```

To read a good comparison of mapping options in R, see this [blog post](https://rstudio-pubs-static.s3.amazonaws.com/324400_69a673183ba449e9af4011b1eeb456b9.html).

We can even view in an interactive mode.

```{r}
tmap_mode("view")   # set tmap mode to interactive
# All together
tm_shape(sle_adm3, bbox = c(-13.3, 8.43, -13.2, 8.5)) +     #
  tm_polygons(col = "#F7F7F7") +
  tm_borders(col = "#000000", lwd = 2) +
  tm_text("admin3name")+
tm_shape(linelist_sf) +
  tm_dots(size=0.2, col='blue', alpha = 1) 
```

Now remember that the variable information for each point on the map is contained in the dataset linelist_sf. We can use this for example to colour each point on a dot based on the outcome recorded (e.g. Death, Recover or No information/Missing).

```{r}
tmap_mode("view")   # set tmap mode to interactive
# All together
tm_shape(sle_adm3, bbox = c(-13.3, 8.43, -13.2, 8.5)) +     #
  tm_polygons(col = "#F7F7F7") +
  tm_borders(col = "#000000", lwd = 2) +
  tm_text("admin3name")+
tm_shape(linelist_sf) +
  tm_dots(size=0.2, col='outcome', alpha = 1) 
```

<!-- ======================================================= -->

## Spatial joins

You may be familiar with *joining* data from one dataset to another one. A spatial join serves a similar purpose but leverages spatial relationships. Instead of relying on common values in columns to correctly match observations, you can utilize their spatial relationships, such as one feature being *within* another, or *the nearest neighbor* to another, or within a *buffer* of a certain radius from another, etc.

The **sf** package offers various methods for spatial joins. See more documentation about the st_join method and spatial join types in this [reference](https://r-spatial.github.io/sf/reference/geos_binary_pred.html)

### Points in polygon 

**Spatial assign administrative units to cases**

Here is an interesting conundrum: the case linelist does not contain any information about the administrative units of the cases. Although it is ideal to collect such information during the initial data collection phase, we can also assign administrative units to individual cases based on their spatial relationships (i.e. point intersects with a polygon).

Below, we will spatially intersect our case locations (points) with the ADM3 boundaries (polygons):

1)  Begin with the linelist (points)\
2)  Spatial join to the boundaries, setting the type of join at "st_intersects"\
3)  Use `select()` to keep only certain of the new administrative boundary columns

```{r, warning=F, message=F}
linelist_adm <- linelist_sf %>%
  # join the administrative boundary file to the linelist, based on spatial intersection
  st_join(sle_adm3)
```

All the columns from `sle_adms` have been added to the linelist! Each case now has columns detailing the administrative levels that it falls within. In this example, we only want to keep two of the new columns (admin level 3), so we `select()` the old column names and just the two additional of interest:

```{r, warning=F, message=F}
linelist_adm <- linelist_sf %>%
  # join the administrative boundary file to the linelist, based on spatial intersection
  st_join(sle_adm3) %>% 
  # Keep the old column names and two new admin ones of interest
  select(names(linelist_sf), admin3name, admin3pcod)
```

Below, just for display purposes you can see the first ten cases and that their admin level 3 (ADM3) jurisdictions that have been attached, based on where the point spatially intersected with the polygon shapes.

```{r, warning=F, message=F}
# Now you will see the ADM3 names attached to each case
linelist_adm %>% select(case_id, admin3name, admin3pcod)
```

Now we can describe our cases by administrative unit - something we were not able to do before the spatial join!

```{r, warning=F, message=F}
# Make new dataframe containing counts of cases by administrative unit
case_adm3 <- linelist_adm %>%          # begin with linelist with new admin cols
  as_tibble() %>%                      # convert to tibble for better display
  group_by(admin3pcod, admin3name) %>% # group by admin unit, both by name and pcode 
  summarise(cases = n()) %>%           # summarize and count rows
  arrange(desc(cases))                     # arrange in descending order

case_adm3
```

We can also create a bar plot of case counts by administrative unit.

In this example, we begin the `ggplot()` with the `linelist_adm`, so that we can apply factor functions like `fct_infreq()` which orders the bars by frequency.

```{r, warning=F, message=F}
ggplot(
    data = linelist_adm,                       # begin with linelist containing admin unit info
    mapping = aes(
      x = fct_rev(fct_infreq(admin3name))))+ # x-axis is admin units, ordered by frequency (reversed)
  geom_bar()+                                # create bars, height is number of rows
  coord_flip()+                              # flip X and Y axes for easier reading of adm units
  theme_classic()+                           # simplify background
  labs(                                      # titles and labels
    x = "Admin level 3",
    y = "Number of cases",
    title = "Number of cases, by adminstative unit",
    caption = "As determined by a spatial join, from 1000 randomly sampled cases from linelist"
  )
```

## Choropleth maps

Choropleth maps can be useful to visualize your data by pre-defined area, usually administrative unit or health area. In outbreak response this can help to target resource allocation for specific areas with high incidence rates, for example.

Now that we have the administrative unit names assigned to all cases (see section on spatial joins, above), we can start mapping the case counts by area (choropleth maps).

Since we also have population data by ADM3, we can add this information to the *case_adm3* table created previously.

We begin with the dataframe created in the previous step `case_adm3`, which is a summary table of each administrative unit and its number of cases.

1)  The population data `sle_adm3_pop` are joined using a `left_join()` from **dplyr** on the basis of common values across column `admin3pcod` in the `case_adm3` dataframe, and column `adm_pcode` in the `sle_adm3_pop` dataframe.
2)  `select()` is applied to the new dataframe, to keep only the useful columns - `total` is total population\
3)  Cases per 10,000 populaton is calculated as a new column with `mutate()`

```{r}
# Add population data and calculate cases per 10K population
case_adm3 <- case_adm3 %>% 
     left_join(sle_adm3_pop,                             # add columns from pop dataset
               by = c("admin3pcod" = "adm3_pcode")) %>%  # join based on common values across these two columns
     select(names(case_adm3), total) %>%                 # keep only important columns, including total population
     mutate(case_10kpop = round(cases/total * 10000, 3)) # make new column with case rate per 10000, rounded to 3 decimals

case_adm3                                                # print to console for viewing
```

Join this table with the ADM3 polygons shapefile for mapping

```{r, warning=F, message=F}
case_adm3_sf <- case_adm3 %>%                 # begin with cases & rate by admin unit
  left_join(sle_adm3, by="admin3pcod") %>%    # join to shapefile data by common column
  select(objectid, admin3pcod,                # keep only certain columns of interest
         admin3name = admin3name.x,           # clean name of one column
         admin2name, admin1name,
         cases, total, case_10kpop,
         geometry) %>%                        # keep geometry so polygons can be plotted
  drop_na(objectid) %>%                       # drop any empty rows
  st_as_sf()                                  # convert to shapefile
```

Now we can map the case incidence, using the blue ramp colour to indicate number of cases.

```{r, message=F, warning=F}
# tmap mode
tmap_mode("plot")               # view static map

# plot polygons
tm_shape(case_adm3_sf) + 
        tm_polygons("cases") +  # color by number of cases column
        tm_text("admin3name")   # name display
```

We can also map the results for the incident rate (per 10,000 people) - let's also change colour to a purple ramp to differentiate this from the number of case map above.

```{r, warning=F, message=F}
# Cases per 10K population
tmap_mode("plot")             # static viewing mode

# plot
tm_shape(case_adm3_sf) +                # plot polygons
  tm_polygons("case_10kpop",            # color by column containing case rate
              breaks=c(0, 10, 50, 100), # define break points for colors
              palette = "Purples"       # use a purple color palette
              ) +
  tm_text("admin3name")                 # display text
```

### Saving our map images

To save our maps - we can do this a few days.

This way describes the `tmap_save()` that we can use from the `tmp` package.

First it can be easier to give our map object in R a 'name', then we can call that object in the `tmap_save()` function.

```{r}
# Cases per 10K population
tmap_mode("plot")             # static viewing mode

# plot
map_case_rate <- tm_shape(case_adm3_sf) +                # plot polygons
  tm_polygons("case_10kpop",            # color by column containing case rate
              breaks=c(0, 10, 50, 100), # define break points for colors
              palette = "Purples"       # use a purple color palette
              ) +
  tm_text("admin3name")                 # display text

tmap_save(map_case_rate, "map_case_rate.png", width=1920, height=1080, asp=0)
```

### 🧠 Discussion Questions

**Take a moment to compare the case count map with the incidence map (cases per 10,000 population).**

Discuss the following:

-   Do the areas with the highest number of cases also have the highest incidence rates?
    -   If they differ, why might that be?
-   How does population size influence the simple case count map?
    -   Does a large number of cases always mean that transmission risk is high?
    -   In what situations might case counts be misleading?
-   What additional information does the incidence map provide?
    -   How might this influence public health decisions such as resource allocation?
-   If you were planning a response, which map would you rely on more — counts or incidence — and why?
-   What limitations or biases should we consider when interpreting these maps?
    -   Case detection
    -   Access to healthcare
    -   Population mobility
    -   Geographic clustering

<!-- ======================================================= -->

## Bonus Exercises

(Complete if you are ahead of time!)

> 🎨 How colour scales & break points change the story a map tells

When creating choropleth maps, the choice of colours and classification scheme (i.e., how values are grouped into categories) can dramatically change how data are interpreted.

In this section, we will experiment with colour palettes and break points to see how the same data can appear very different depending on choices made by the analyst.

**Why does this matter?**

-   Human perception is strongly influenced by colour contrast.
-   Changing break points can make a pattern look more extreme or more subtle.
-   Different classification methods (equal intervals, quantiles, manual breaks) highlight different features.
-   Public health decisions may accidentally be influenced by how a map is styled.
-   Transparency about methods is essential for responsible map‑making.

### Exercise 1: Experiment with case counts

We will repeat the map of number of cases but apply different palettes and break structures. For each map you generate, ask:

-   Does the map look more “alarming” or more “calm”?
-   Which areas appear to be hotspots? Are they the same areas in each version?
-   How does the number of breaks influence interpretability?
-   Could this style accidentally mislead someone?

```{r, message=F, warning=F}
# tmap mode
tmap_mode("plot")               # view static map

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("cases") + # leave without any parameters and let colours and scale autoselect
  tm_text("admin3name")

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("cases", palette = "YlOrBr") + # now we are setting the colour palette to a particular scheme
  tm_text("admin3name")

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("cases",
              palette = "YlOrBr",
              breaks = c(0, 10, 50, 100, 200, 300)) + # let's also add in some specified break points for the colour palette
  tm_text("admin3name")

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("cases",
              palette = "YlOrBr",
              breaks = c(0, 10, 50, 100, 150, 200, 250, 300)) + # what about adding even more specified break points for the colour palette?
  tm_text("admin3name")
```

### Exercise 2: Experiment with incidence number

Now perform the same experiment for the incidence map. Reflect on:

-   Whether colour choices influence your perception more strongly for raw counts or for rates
-   Whether small population areas appear artificially “high risk”
-   How break selection can obscure or exaggerate true epidemiological patterns

```{r, message=F, warning=F}
# tmap mode
tmap_mode("plot")               # view static map

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("case_10kpop") + # leave without any parameters and let colours and scale autoselect
  tm_text("admin3name")

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("case_10kpop", palette = "YlOrBr") + # now we are setting the colour palette to a particular scheme
  tm_text("admin3name")

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons("case_10kpop",
              palette = "YlOrBr",
              breaks = c(0, 10, 50, 100, 200, 300)) + # let's also add in some specified break points for the colour palette
  tm_text("admin3name")

# plot polygons
tm_shape(case_adm3_sf) +
  tm_polygons(
    "case_10kpop",
    palette = "YlOrBr",
    breaks = c(0, 10, 50, 100, 150, 200, 250, 300)
  ) + # what about adding even more specified break points for the colour palette?
  tm_text("admin3name")   
```

### Summary Reflection

After trying a few colour schemes:

-   Which palette feels most appropriate for public health communication?
-   Which break method best reflects the epidemiological story?
-   How might colour choice affect political or community reactions?
-   What principles would you follow when publishing a map?

::: {.callout-note icon="false"}
Maps are analytical tools but also communication tools — and each choice we make influences interpretation.
:::
